---
title: "Regression Models Course Project"
author: "J. Thatcher"
date: "Tuesday, January 19, 2016"
output: word_document
---
#1.0 Introduction
In this project we consider the mtcars dataset. We will explore the relationship between transmission type (automatic or manual) and miles per gallon (MPG).

#2.0 Data Analysis
###2.1 Exploratory Data Analysis
In `data(mtcars)` we find that the first column variable is `mpg` (Miles/US Gallon) and the nineth column variable `am` is transmission type (0 = automatic, 1 = manual)


```{r warning=F, message=F}
library(ggplot2)
DF <- data.frame(mtcars) #grab data as data.frame
```

Fist, an exploratoty graph of the response and predictor using `ggplot2`.

```{r fig.height=3, fig.width=4}
#plot the data and a linear fit
ggplot(DF, aes(x = am, y = mpg)) + 
        geom_point(alpha = 0.5, cex = 3) + 
        geom_smooth(method = "lm") +
        ylab(label = "hello") +
        xlab(label = "hi")
```

The graph shows that there may be a significant correlation between mpg and transmission type.

###2.2 Relationship Between MPG and Transmission Type
Initially we will ask: is MPG better for manual or automatic transmission vehicles?

let's explore this using a linear model. First, we will calculate the coefficients of a linear model using the `lm` function.

```{r}
#fit a linear model
fit <- lm(mpg ~ am, data = DF)
summary(fit)
```

###2.3 Interpretation of Correlation Coefficient, R-Squared
Reviewing the calculations performed by the `lm` funciton, we can find many imporant values including the correlation coefficient. The corelation coefficeint is a measure of the variance in vehicle MPG that is explained by the variable transmission type. The R-squared value for the model is 0.36. Which means that transmission type explains 36% of the variance in vehicle MPG.

###2.4 Interpretation of the Intercept, Beta0
Following these calculations, the y-intercept Beta0 = 17.15. This values is the mean mpg for the automatic transmission vehicles.  

###2.5 Interpretation of the Slope, Beta1
Reviewing the slope, Beta1, in the model will answer our initial quesiton, is MPG better for manual or automatic transmission vehicles? Beta1 is 7.25, and is interpreted as the change in the average vehicle's MPG for one unit increase in the predictor variable, transmission. Because Transmission Type is a factor variable, Beta1 + Beta0 = the mean MPG for manual transmission vehicles. Beta1 is positive, which means that there is an increase in mpg for manual transmission vehicles. Furthermore, the difference in MPG between automatic and manual transmission vehicles is 7.25 mpg.

###2.6 Is the slope significant?
We can perform a hypothesis test on the Beta1 coefficient to determine that the slope is actually significant. The hypotheses are:

H0: Beta1  = 0 
H1: Beta1 != 17.2

To determine whether we can reject the null hypothesis, H0, we will go back to our results from the `lm` function. In the second row and fourth column we find the probability that the the slope is equal to zero equals 0.000285. This is much lower than our level of statistical significance, alpha = 0.05. Therefore, we can reject H0 and determine that Beta1 is significantly different that zero.

###2.7 Residuals
An important way to determine the effectiveness of the linear model is to plot the residuals. In a residual plot, we are looking for the residual points to be spread in a random fashion on either side of the horizantle line. This will help confirm that the linear model was an appropriate model for this data.

```{r fig.height=3, fig.width=4}
#plot the residuals
ggplot(fit, aes(x = DF$am, y = resid(fit))) + 
        geom_point(alpha = 0.5, cex = 3) + 
        geom_hline(yintercept=0, color="red") +
        ylab(label = "hello") +
        xlab(label = "hi")
```

The plot does show that the residuals are spread evenly above and below the line at 0. However, the graph shows that the data slightly resembles a property called Heteroscedasticity. This is when the spread of the data differs at different levels of the predictor variable. In this case the variance of mpg appears to be slightly higher in cars with manual transmissions compared to automatic transmissions.

###2.8 Alternative Models
Upon reviewing the residual plot, we will consider a few other models that may show better performance compared to the linear model.  

```{r fig.height=4, fig.width=6}

```


#3.0 Executive Summary
The results of this analysis answer the two main questions asked. One, "Is an automatic or manual transmission better for MPG?" Answer, the manual transmission is on average better for MPG. Two, "Quantify the MPG difference between automatic and manual transmissions." Answer: using a linear regression model we find that there is an average increase of 7.25 mpg for cars with a manual transmission.